---
title: "hw1"
author: "Matija Jakovac, Dunja Petrović, Leo Sérra"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(RColorBrewer)
library(ggplot2)
library(reshape2)
library(viridis)
library(circlize)
library(gridExtra)
library(ggridges)
library(knitr)
```

```{r}
df <- read.csv(file = "euroleague_23_24.csv", header = TRUE, sep = ";", dec=",")
head(df)
```

# 1. First do the exploratory data analysis.
## a) Discard the variable “No” from the data set. (1p)

```{r}
df <- select(df, -No)
df
```
## b) Split variable “Min” using strsplit() function. Give the name “aux” to the output. The first element of each row will show the minutes that the player played in total. (1p)

```{r}
aux <- strsplit(as.character(df$Min), split = ":")
head(aux)
```
## c) Add a numerical variable to the data set named “Min 2” which shows on average how many minutes each player played in the game. (2p)
```{r}
min2 <- sapply(aux, function(x) {
  minutes <- as.numeric(x[1])
  seconds <- as.numeric(x[2])
  total_time <- minutes + seconds / 60
  return(total_time)
})

df$Min2 <- min2
df$Min <- NULL

head(df)
```
## d) Check the structure of the data and assign correct type to each variable considering whether it is a categorical or numerical variable. (2p)
```{r}
str(df)
df$TEAM <- as.factor(df$TEAM)
df$PLAYER <- as.factor(df$PLAYER)
df$POSITION <- as.factor(df$POSITION)

df$GP <- as.numeric(df$GP)
df$GS <- as.numeric(df$GS)
df$PTS <- as.numeric(df$PTS)
df$X2P. <- as.numeric(df$X2P.)
df$X3P. <- as.numeric(df$X3P.)
df$FT. <- as.numeric(df$FT.)
df$OR <- as.numeric(df$OR)
df$DR <- as.numeric(df$DR)
df$TR <- as.numeric(df$TR)
df$AST <- as.numeric(df$AST)
df$STL <- as.numeric(df$STL)
df$TO <- as.numeric(df$TO)
df$BLK <- as.numeric(df$BLK)
df$BLKA <- as.numeric(df$BLKA)
df$FC <- as.numeric(df$FC)
df$FD <- as.numeric(df$FD)
df$PIR <- as.numeric(df$PIR)
df$Min2 <- as.numeric(df$Min2)

head(df)
```
#2. Application of PCA.

## a) Apply PCA on all the scaled numerical variables in the data set by using PCA() function in FactoMineR package. Treat the categorical variables and the variable “PIR” as suplemantary variables using arguments quali.sup and quanti.sup correctly. (3p)
```{r}
library(FactoMineR)

names(df)

dffa <- PCA(df, scale.unit = TRUE, ncp = 5, quali.sup = c(1, 2, 3), quanti.sup = 13, graph = TRUE)
```
```{r}
summary(dffa)
```

## b) How many components should be extracted? Decide on the number of components considering eigenvalues. (3p)
```{r}
eigenvalues <- dffa$eig
eigenvalues

which(eigenvalues[, "eigenvalue"] > 1)
```
According to the Kaiser criterion (eigenvalue > 1), the first 4 components should be extracted.

## c) Interpret the loadings/correlations of variables at each dimension (3p).
```{r}
dffa$var$cor
```
Dimension 1 --> strongest positive correlations with GP and Min2...strong positive correlations with GS, PTS, FT., DR, TR, STL, TO and FD
Dimension 2 --> strong positive correlation with FT....strong negative correlation with GS (strongest correlation overall)
Dimension 3 --> very strong positive correlation with X3P., strong positive correlation with PIR
Dimension 4 --> no particularly strong positive or negative correlations (-0.48 negative correlation with DR)
Dimension 5 --> no particularly strong positive or negative correlations (0.46 positive correlation with X2P.)

## d) Use plot.PCA() function to show correlations between variables and the extracted dimensions. (For the variables you should use the argument choix = “var”). Plot all the extracted dimensions changing argument “axes”.(3p)
```{r}
plot(dffa, choix= "var", axes = c(1, 2), main="Correlation between Variables and PC1 and PC2", cex=0.6)
plot(dffa, choix= "var", axes = c(1, 3), main="Correlation between Variables and PC1 and PC3", cex=0.6)
#plot(dffa, choix= "var", axes = c(1, 4), main="Correlation between Variables and PC1 and PC4", cex=0.6)
plot(dffa, choix= "var", axes = c(2, 3), main="Correlation between Variables and PC2 and PC3", cex=0.6)
#plot(dffa, choix= "var", axes = c(2, 4), main="Correlation between Variables and PC2 and PC4", cex=0.6)
#plot(dffa, choix= "var", axes = c(3, 4), main="Correlation between Variables and PC3 and PC4", cex=0.6)
```

## e) Interpret variable plots. How can each dimension be named?

**Dimension 1 (Usage/Production volume)**
- pointing strongly to the positive side are GP, GS, Min2, PTS, TR/DR, PIR, TO, FD and mainly all variables except maybe X3P. which smallest positive correlation with the dimension
- players that do appear more on that side are the ones who play a lot and contributed a lot to the team having high box-score numbers such as points, rebounds, fouls and turnovers which happen more often playing more minutes
- low-values of those numbers appers to the bench players

**Dimension 2 (Perimeter vs Interior)**
- having positive correlation with the dimension are X3P., AST and STL while negative correlation are for variable OR, BLK and TR
- this correlations represent separation between perimeter play with outside shooting and play-making that are often characteristics of guards versus presence in the paint with rebounding and blocks often connected with centers

**Dimension 3 (Finishing efficiency)**
- X2P. and FT have the most positive correlation with dimension while FD, PIR and BLKA have the most negative ones
- having higher 2-point conversion percentage as well as free throw percentage show that player has high efficiency while on the other side if player gets blocked (BLKA) his efficiency will drop

## f) Show individual pilots for the extracted dimensions changing argumennt choix=“ind” in plot.PCA() function.

```{r}
fsize <- 0.45

plot(dffa, choix = "ind", axes = c(1, 2),
     label = "all",
     title = "Individuals – PC1 vs PC2", cex=fsize)

plot(dffa, choix = "ind", axes = c(1, 3),
     label = "all", cex=fsize,
     main  = "Individuals – PC1 vs PC3")

plot(dffa, choix = "ind", axes = c(2, 3),
     label = "all", cex=fsize,
     main  = "Individuals – PC2 vs PC3")

```
## g) Interpret the individual plots.

The individual plots show the same correlations as the variable plots but examining rather individuals players.

PC 1 vs PC2

High PC1 and PC2 -> we see high-usage and perimeter oriented players mostly guards that play a lot of minutes, shoot threes and playmake. The examples are Kendrick Nunn, Facundo Campazzo, Kostas Sloukas, Thomas Walkup.

High PC1 and low PC2 -> high-usage players but playing in the interior. Those type of players are starting centers who rebound, block and draw fouls. The examples are Mathias Lessort and Walter Tavares.
Low PC1 and high PC2 -> low minute perimeters reserves such as Alexandros Samodurov and Dimitris Moraitis.
Low PC1 and PC2 -> low minute interior players such as Ismaila Diagme and Georgios Tsalmpouris.

PC1 vs PC3

High PC3 -> players that are efficient finishers that are rarely blocked such as Hugo Gonzalez and Panagiotis Kalaitzakis.
Low PC3  -> are the ones with lower percentages and more shots blocked such as Mathias Lessort despite high usage and Tyler Dorsey.

PC2 vs PC3

High PC2 and PC3 -> efficient and perimeter skills - Rudy Fernandez
High PC2 and low PC3 -> not that efficient guards - Kendrick Nunn
Low PC2 and high PC3 -> efficient interior players - Vincent Poirier
Low PC2 and low PC3 -> not that efficient interior players - Mathias Lessort


# 3. Application of MDS

## a) Apply metric MDS using Euclidean distance on scaled numerical variables.

```{r}
euro_num  <- df |> select(where(is.numeric)) 
d_euro <- dist(scale(euro_num), method = "euclidean")
mds_euro <- cmdscale(d_euro, k = 2, eig = TRUE)
cat("GOF of MDS:")
mds_euro$GOF
```

## b) Plot the data using the points on the first two coordinates using players names as label.

```{r}
players <- df$PLAYER
x <- mds_euro$points[,1]
y <- mds_euro$points[,2]
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2",
     main="Metric MDS", type="n")
text(x, y, labels = players, cex=0.6)
```

## c) Interpret the plot.

We can only make clusters having similar characteristics interpreting this plot.

We can see in the bottom right players such as M. Fall, N. Milutinov, V. Poirier, W. Tavares and M. Lessort have similar characteristics. We can diagonal border/gap between them and other players

In the top left we can say K. Nunn and F. Campazzo have also similar characteristics.

In the top right we can extract A. Samodurov and D. Moraitis which are similar.

In general players which labels overlap (K. Antetokounmpo and A. Noua) or in the same part of plot have similar characteristics. Comparing/combining those clusters with PCA results we can come to the conclusion which are does characteristics. 


